Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-08-31 manu-draft.qmd
working on opportunity and restoration area results text, added summary objects for areas

load(file = here('tabs/subttab.RData'))
load(file = here('tabs/acretab.RData'))
load(file = here('tabs/currtab.RData'))
load(file = here('tabs/targtab.RData'))
```

@@

### Restoration database

Past restoration projects were quantified for each of the major habitat types, when available, to inform expectations for setting short-term goals and long-term targets. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single database. Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program. The collected data included project name, year, description, size (area or length), and location (latitude and longitude). Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals. Living shoreline projects, including seawall enhancements and oyster reef installations, were inventoried separately.

### Creating and combining opportunity layers

@@

Current estimates for subtidal habitats without historical trend information included 171 ha for natural hard bottom and 67 hectares for artificial reefs.  The former estimate was obtained from previous mapping work to locate and quantify these habitats using sonar mapping [@CSA19].  The latter estimate was obtained from county governments that manage artificial reefs in the bay.

Current estimates and trend information were available for seagrasses, tidal flats, and oyster bars (Table 1). Oyster bars were estimated at 67 ha in 2018 (Table 1), showing a 29% increase since mapping began in 2014.  This increase may represent improved ground-truthing and photointerpretation of oyster bars from aerial photography. Tidal flats have generally increased in area from 1988 to the mid-2000s, followed by a decrease to 2018.  The current estimate for tidal flats is 6,569 hectares, showing a 24% decline compared to the 1988 estimate of 8,700 hectares. Seagrasses have increased by 75% since 1988 to a current estimate of 6,986 ha. The change analysis comparing 1988 to 2018 for subtidal habitats (@fig-subtchgdatalluv) confirmed trends in Table 1 and showed that the seagrass increases were primarily associated with the colonization of non-vegetated areas of tidal flats, as well as unclassified areas of open water.

Total intertidal habitat (mangroves, salt barrens, and salt marshes) showed a 12% increase of 835 ha from 1990 to 2017, with a current estimate of 8,340 ha (Table 2). Mangroves increased by 15% to 6,276 ha, salt barrens increased by 7% to 203 ha, and salt marshes increased by 3% to 1,861 ha. The current estimate of the extent of mangrove forests in Tampa Bay is 6,192 ha. Mangrove forest coverage increased by 684 ha (12%). Salt marshes in Tampa Bay cover 1,844 ha, and coverage increased by 30 ha (2%). Despite a net increase in salt marsh habitat, the change analysis showed that 153 ha were changed to mangrove habitat (@fig-chgdatalluv). The current extent of tidal tributary length is 622 km (no trend information is available).

Trend assessments for supratidal habitats showed the effects of increasing land development in the Tampa Bay watershed (Table 2).  Developed lands showed a 44% increase to 217,047 ha from 1990 to 2017, with a notable 30% decrease to 1,446 ha for coastal uplands, a 38% decrease to 57,836 ha for native uplands, and an 18% decrease to 189,512 ha for restorable lands. Non-forested freshwater wetlands showed a 24% increase to 27,358 ha, while forested freshwater wetlands decrased by 5% to 61,667 ha.  The change analysis (@fig-chgdatalluv) showed that a majority of land conversion to developed came from restorable lands (21,292 ha) and native uplands (7,184 ha), with smaller proportions converted from forested freshwater wetlands (1,407 ha) and coastal uplands (193 ha). Supratidal habitats converted to restorable primarily included native uplands (8,304 ha), forested freshwater wetlands (1,700 ha), and developed (2,794 ha).  The increase in non-forested freshwater wetlands was primarily from restorable lands (2,759 ha).

A total of 460 projects were documented between 1971 and 2019 (Table 4), addressing the full range of habitat types, including Estuarine (n=228), Freshwater (n=53), Uplands (n=119), and Mixed (n=60). A total of 1,978 ha have been restored, and 12,930 ha and 42.8 km of linear projects were enhanced during the time period. Forty lead partners were documented as responsible for the projects, although some of these lead partners are departments within the same overall agency. Eighty-nine living shoreline projects, seawall enhancements, and oyster reef module installations along shorelines were inventoried, with a linear footprint of 18.2 km.

### Habitat restoration opportunities

@fig-oppmap

In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable. Developed areas in the Tampa Bay watershed encompassed 2,172 km^2^ (36.5%) of the 5,872 km^2^ watershed area. Between 2014 and 2017, the developed footprint increased by seven percent (7%, Fig 2).

@@

Xeric soils were roughly aggregated in an east-west band through the middle of the watershed (Fig 4). Approximately 2,107 km2 was classified as xeric, while 3,760 km2 was classified as mesic/hydric.

>>>>

load(file = here('tabs/subttab.RData'))
load(file = here('tabs/acretab.RData'))
load(file = here('tabs/currtab.RData'))
load(file = here('tabs/targtab.RData'))
load(file = here('data/areas.RData'))
```

@@

### Restoration and enhancement projects

Past restoration projects were quantified for each of the major habitat types, when available, to inform expectations for setting short-term goals and long-term targets. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years was compiled from various sources.  Restoration was defined as any activity that involved earthwork to reshape the land or the addition of structural elements (e.g., rock). Enhancement was defined as any activity not including earthwork that improved the environment (e.g., planting native vegetation, invasive species or debris removal, prescribed burns, etc.). Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program. The collected data included project name, year, description, size (area or length), and location (latitude and longitude). Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals. Living shoreline projects, including seawall enhancements and oyster reef installations, were inventoried separately.

### Opportunity areas and restoration potential

@@

Current estimates and trend information were available for seagrasses, tidal flats, and oyster bars (Table 1). Oyster bars were estimated at 67 ha in 2018 (Table 1), showing a 29% increase since mapping began in 2014.  This increase may represent improved ground-truthing and photointerpretation of oyster bars from aerial photography. Tidal flats have generally increased from 1988 to the mid-2000s, followed by a decrease to present.  The current estimate for tidal flats is 6,569 hectares, showing a 24% decline compared to the 1988 estimate of 8,700 hectares. Seagrasses have increased by 75% since 1988 to a current estimate of 6,986 ha. The change analysis comparing 1988 to 2018 for subtidal habitats (@fig-subtchgdatalluv) confirmed trends in Table 1 and showed that the seagrass increases were primarily associated with the colonization of non-vegetated areas of tidal flats, as well as unclassified areas of open water. Current estimates for subtidal habitats without historical trend information included 171 ha for natural hard bottom and 67 hectares for artificial reefs.

Total intertidal habitat (mangroves, salt barrens, and salt marshes) showed a 12% increase of 835 ha from 1990 to 2017, with a current estimate of 8,340 ha (Table 2). Mangroves increased by 15% to 6,276 ha, salt barrens increased by 7% to 203 ha, and salt marshes increased by 3% to 1,861 ha. Despite a net increase in salt marsh habitat, the change analysis showed that 153 ha were replaced by mangrove habitat (@fig-chgdatalluv). The current extent of tidal tributary length is 622 km (no trend information is available).

Trend assessments for supratidal habitats showed the effects of increasing land development and loss of restorable habitats in the Tampa Bay watershed (Table 2).  Developed lands showed a 44% increase to 217,047 ha from 1990 to 2017, with a notable 30% decrease to 1,446 ha for coastal uplands, a 38% decrease to 57,836 ha for native uplands, and an 18% decrease to 189,512 ha for restorable lands. Non-forested freshwater wetlands showed a 24% increase to 27,358 ha, while forested freshwater wetlands decreased by 5% to 61,667 ha.  The change analysis (@fig-chgdatalluv) showed that a majority of land conversion to developed came from restorable lands (21,292 ha) and native uplands (7,184 ha), with smaller proportions converted from forested freshwater wetlands (1,407 ha) and coastal uplands (193 ha). Habitats converted to restorable primarily included native uplands (8,304 ha), forested freshwater wetlands (1,700 ha), and developed (2,794 ha).  The increase in non-forested freshwater wetlands was primarily from restorable lands (2,759 ha).

### Habitat restoration and enhancement

A total of 460 restoration projects were documented in Tampa Bay and its watershed between 1971 and 2019.  These projects were divided among habitat types that included estuarine (n = 228), freshwater (n = 53), uplands (n = 119), and a mix of all three (n = 60). A total of 1,978 ha have been restored, whereas 12,930 ha and 42.8 km (as shoreline or tributaries) were enhanced. Forty partners were responsible for these projects, although some were from departments within the same agency. Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, with a total linear length of 18.2 km.

### Summary of opportunity areas and restoration potential

The geospatial analysis that produced a map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap). Native lands categorized as existing conservation native, proposed conservation native, or reservation native totaled.  Similarly, restorable lands categorized as existing conservation restorable, proposed conservation restorable, or reservation restorable totalled.  Understandably, most of the native and restorable lands occurred in undeveloped areas of the watershed, particularly in the north and southwest. Existing conservation lands (existing conservations native, existing conservation restorable) totalled and proposed conservation lands (proposed conservation native, proposed conservation restorable).  Areas in the coastal stratum (the MLLW to 5-foot contour) included of reservation native habitats and ha of reservation restorable habitats.

Combining the restorable lands in @fig-oppmap with the coastal stratum, soils, and salinity data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).  The restortation potential
In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable. Developed areas in the Tampa Bay watershed encompassed 2,172 km^2^ (36.5%) of the 5,872 km^2^ watershed area. Between 2014 and 2017, the developed footprint increased by seven percent (7%, Fig 2).

@@
